---
title: Data Analysis
author: Thomas E. Saunders
---
## Set up

Load libraries

```{r}
#| message: false

library(akldogs)
library(tidyverse)
library(tidytext)
library(fs)
library(gt)
library(patchwork)
library(scales)
```

Read in data compiled from Animal Management reports

```{r}
#| message: false

report_files <- list.files(path = "../data", 
                           pattern = "am-reports", 
                           full.names = TRUE)

report_data <- 
  report_files |> 
  map(read_csv) |> 
  set_names(path_file(str_extract(report_files, "[^-]+(?=\\.[^.]+$)")) |> path_ext_remove())
```

Tidy data and column names

```{r}
registration <- 
  registration |> 
  mutate(
    across(
      starts_with("animal_breed"), 
      ~ str_replace(.x, "^([^,]+),\\s*(.*)$", "\\2 \\1")
    )
  )

registration <- rename(registration, suburb = suburb_name)
rfs <- rename(rfs, suburb = location_suburb_name)
```

## Auckland Dog Population Overview

Known dogs as at FY2025

```{r}
reg_2025 <-
  registration |> 
  filter(sheet_name == "FY25")

nrow(reg_2025)
```

Dog population primary breed counts

```{r}
primary_breed <-
  reg_2025 |> 
  group_by(animal_breed_description) |> 
  count(name = "primary_breed") |> 
  ungroup() |> 
  mutate(
    prop = primary_breed / sum(primary_breed)
  ) |> 
  arrange(desc(prop))
```

Dog population secondary breed counts

```{r}
secondary_breed <-
  reg_2025 |> 
  group_by(animal_breed2_description) |> 
  count(name = "secondary_breed") |> 
  ungroup() |> 
  mutate(
    prop = secondary_breed / sum(secondary_breed)
  ) |> 
  arrange(desc(prop))
```

Primary breeds most commonly classified as 'menacing' due to behaviour

```{r}
menacing_breed <-
  reg_2025 |>
  filter(classification == "Menacing - observed or reported behaviour of dog") |> 
  count(animal_breed_description, name = "menacing") |> 
  left_join(primary_breed |> rename(primary_prop = prop), by = join_by(animal_breed_description)) |> 
  mutate(menacing_prop = menacing / primary_breed) |> 
  arrange(desc(menacing_prop))
```

Annual attack rate

```{r}
attack_rate <-
  report_data$service |> 
  filter(category == "attack-people") |>
  select(-category) |> 
  rename(count_attacks = count) |> 
  left_join(
    report_data$ownership |> 
      filter(category == "dogs") |> 
      select(-category) |> 
      rename(count_dogs = count),
    by = join_by(year)) |> 
  mutate(
    attack_rate = (count_attacks / count_dogs) * 1000,
  ) 
```

Severity of attack based on primary dog breed

```{r}
attack_severity <-
  impounds |> 
  filter(impound_date > "2023-07-01" & !is.na(severity)) |> 
  group_by(severity) |> 
  count(primary_breed) |> 
  arrange(desc(severity), desc(n))
```

Trends in ownership over time

```{r}
ownership_trends <-
  report_data$ownership |>
  pivot_wider(
    names_from = category,
    values_from = count
  ) |> 
  arrange(year) |>
  mutate(
    prev_value = lag(dogs),
    change = dogs - prev_value,
    pc_change = change / dogs,
    reg_pc = registered / dogs,
    desex_pc = desexed / dogs,
    men_pc = menacing / dogs
  ) |>
  select(year, dogs, pc_change:men_pc) |> 
  arrange(desc(year))

ownership_trends |>
  rename("Year" = year,
         "Known Dogs" = dogs,
         "Pop. Change (%)" = pc_change,
         "Registered (%)" = reg_pc,
         "Desexed (%)" = desex_pc,
         "Menacing (%)" = men_pc) |> 
  arrange(desc(`Year`)) |> 
  gt() |> 
  fmt_percent(columns = contains("%"),
              decimals = 1) |>
  fmt_number(columns = `Known Dogs`,
             sep_mark = ",",
             decimals = 0) |> 
  sub_missing(
    columns = `Desexed (%)`,
    missing_text = "-"
  )
```

Impound outcomes

```{r}
#| warning: FALSE

impound_result <-
  report_data$impounds |> 
  filter(category != "impounds")

impound_result$category <- factor(impound_result$category, 
                                 levels = c("transferred", "returned", "euthanised"))

ggplot(impound_result, aes(x = year, y = count, fill = category)) +
  geom_bar(position = "fill", stat = "identity") +
  theme_minimal() +
  scale_fill_manual(values = c("lightblue", "lightgreen", "red")) +
  scale_y_continuous(labels = scales::label_percent()) +
  scale_x_continuous(breaks = unique(impound_result$year)) +
  xlab("") +
  ylab("")
```

Impound outcome by primary breed (FY2025 population data with FY2024 impound data)

```{r}
breed_processed <- 
  impounds |>
  filter(impound_date > "2023-07-01", status %in% c("Euthanised", "For Euthanasia")) |>
  mutate(primary_breed = if_else(str_detect(primary_breed, "Pit Bull Terrier"), "American Pit Bull Terrier", primary_breed)) |>
  count(primary_breed, name = "count_euthanised") |>
  mutate(proportion_euthanised = count_euthanised / sum(count_euthanised)) |>
  left_join(primary_breed |> rename(count_population = primary_breed,
                                    proportion_population = prop), 
            by = join_by(primary_breed == animal_breed_description)) |>
  mutate(
    diff = proportion_euthanised - proportion_population,
    max_val = pmax(proportion_euthanised, proportion_population)
  ) |>
  filter(abs(diff) > 0.01) |>
  arrange(abs(diff)) |>
  mutate(primary_breed = fct_inorder(primary_breed))

breed_long <- 
  breed_processed |>
  pivot_longer(cols = starts_with("proportion"), names_to = "type", values_to = "val") |>
  mutate(label = if_else(str_detect(type, "pop"), "Population", "Euthanised"))

colours <- c("proportion_euthanised" = "#BF2F24", "proportion_population" = "#436685")

n_breeds <- nrow(breed_processed)

plot_main <- 
  ggplot(breed_long, aes(val, primary_breed)) +
  geom_line(aes(group = primary_breed), color = "#E7E7E7", linewidth = 3.5) +
  geom_point(aes(color = type), size = 3) +
  geom_text(aes(label = percent(val, accuracy = 0.01), color = type,
                hjust = if_else(val == max_val, -0.2, 1.2)), 
            size = 3.25) +
  geom_text(data = filter(breed_long, diff == max(diff)),
            aes(label = label, color = type), 
            nudge_y = 0.8, fontface = "bold", size = 3.25) +
  scale_color_manual(values = colours) +
  coord_cartesian(xlim = c(-0.02, 0.45), ylim = c(1, n_breeds + 1), clip = "off") +
  labs(title = "Breed Profile: Auckland Dog Population vs Euthanised",
       subtitle = "Breeds with >1% difference", x = NULL, y = NULL) +
  theme_minimal() +
  theme(legend.position = "none", panel.grid = element_blank(),
        plot.title = element_text(face = "bold", hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))

plot_diff <- 
  ggplot(breed_processed, aes(x = 0, y = primary_breed)) +
  geom_text(aes(label = percent(diff, accuracy = 0.1), 
                color = if_else(diff > 0, "proportion_euthanised", "proportion_population")),
            fontface = "bold", size = 3.25) +
  annotate("text", x = 0, y = n_breeds + 0.8, 
           label = "Difference", fontface = "bold", size = 3.25) +
  scale_color_manual(values = colours) +
  coord_cartesian(ylim = c(1, n_breeds + 1), clip = "off") +
  theme_void() +
  theme(legend.position = "none")

plot_main + plot_diff + 
  plot_layout(widths = c(10, 1)) & 
  theme(plot.margin = margin(t = 20, r = 5, b = 10, l = 5))

```

Reasons for euthanasia

```{r}
euthanised_reason <-
  impounds |>
  filter(impound_date > "2023-07-01" & !is.na(euthanasia_reason)) |>
  mutate(
    euthanasia_reason = case_when(
      str_detect(euthanasia_reason, "Health|Parvo") ~ "Health or Parvo Virus",
      str_detect(euthanasia_reason, "Dog vs Dog|Dog vs Domestic|Dog vs Stock|Dog vs Poultry") ~ "S.57 Dog vs Animal",
      .default = euthanasia_reason
    )
  ) |> 
  group_by(euthanasia_reason) |> 
  count(name = "count") |>
  ungroup() |> 
  mutate(
    proportion = count / sum(count)
  ) |> 
  arrange(desc(proportion))
```

## Auckland Dog Population By Region

Known dogs

```{r}
regions_dogs <- 
  registration |>
  filter(sheet_name == "FY25") |> 
  left_join(locality, by = join_by(suburb)) |>
  select(suburb, local_board, region) |> 
  group_by(suburb) |> 
  add_count(suburb, name = "known_suburb") |> 
  group_by(local_board) |> 
  add_count(local_board, name = "known_board") |> 
  group_by(region) |> 
  add_count(region, name = "known_region") |>
  unique() |> 
  arrange(desc(known_suburb)) |> 
  ungroup()
```

Registration data

```{r}
regions_2025 <-
  registration |> 
  filter(sheet_name == "FY25") |> 
  mutate(
    current_reg = case_when(registration_latest_year < 2024 ~ FALSE, .default = TRUE),
    owner_reg = case_when(owner_registration_class == "RDOL" ~ TRUE),
    classification = case_when(classification == "Unknown" ~ FALSE,
                               str_detect(classification, "Menacing|Dangerous") ~ TRUE)
  ) |> 
  left_join(regions_dogs, by = join_by(suburb)) |> 
  select(current_reg, owner_reg, animal_breed_description, animal_desexed, classification, suburb, local_board:known_region) 
  
regions_reg <- 
  regions_2025 |>
  pivot_longer(
    cols = c(current_reg, owner_reg, animal_desexed, classification),
    names_to = "metric",
    values_to = "is_true"
  ) |>
  group_by(suburb, local_board, region, metric) |>
  summarise(
    count = sum(is_true, na.rm = TRUE),
    known_suburb = first(known_suburb),
    known_board = first(known_board),
    known_region = first(known_region),
    .groups = "drop"
  ) |>
  group_by(metric) |>
  mutate(
    p_suburb = count / known_suburb
  ) |>
  group_by(local_board, metric) |>
  mutate(p_board = sum(count) / first(known_board)) |>
  group_by(region, metric) |>
  mutate(p_region = sum(count) / first(known_region)) |>
  ungroup()

write_csv(regions_reg, "../data/regions_reg.csv")
```

Compare Local Boards

```{r}
plot_data <- regions_reg |>
  select(local_board, metric, p_board) |>
  distinct()

ggplot(plot_data, aes(x = p_board, y = fct_reorder(local_board, p_board), fill = metric)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~metric, scales = "free_x") +
  scale_x_continuous(labels = scales::label_percent()) +
  theme_minimal() +
  labs(
    title = "Animal Registration Metrics by Local Board (FY25)",
    subtitle = "Proportions calculated against known totals per board",
    x = "Proportion (%)",
    y = NULL
  )
```

Attacks per 1,000 known dogs in each Local Board

```{r}
regions_attacks <-
  rfs |> 
  filter(rfs_date >= "2024-07-01",
         rfs_type %in% c("Dog attack on people", 
                         "Dog attack on child", 
                         "Dog Attack on Child")
  ) |> 
  left_join(regions_dogs |> select(-local_board), by = join_by(suburb)) |> 
  group_by(suburb) |> 
  add_count(suburb, name = "attacks_suburb") |> 
  group_by(local_board) |> 
  mutate(
    attacks_board = sum(attacks_suburb)
  ) |> 
  group_by(region) |> 
  mutate(
    attacks_region = sum(attacks_suburb)
  ) |> 
  ungroup() |> 
  mutate(
    rate_suburb = (attacks_suburb / known_suburb) * 1000,
    rate_board = (attacks_board / known_board) * 1000,
    rate_region = (attacks_region / known_region) * 1000
  ) |> 
  select(suburb, local_board, region:rate_region) |> 
  unique()

write_csv(regions_attacks, "../data/regions_attacks.csv")
```

Primary breed profile for each region

```{r}
regions_breed <-
  registration |> 
  filter(sheet_name == "FY25") |> 
  left_join(locality, by = join_by(suburb)) |> 
  group_by(region) |> 
  count(animal_breed_description, name = "count") |> 
  arrange(desc(count), .by_group = TRUE) |> 
  left_join(regions_dogs |> select(region, known_region) |> unique(), by = join_by(region)) |>
  mutate(
    prop_breed = format(count / known_region, scientific = FALSE)
  )
```

Compare regional breed profiles statistically

```{r}
# Chi-squared test of independence and Pearson residual analysis

# Prepare contingency table
breed_table <- 
  regions_breed |> 
  filter(!is.na(region), !is.na(animal_breed_description)) |> 
  select(region, animal_breed_description, count) |> 
  pivot_wider(names_from = region, values_from = count, values_fill = 0) |> 
  column_to_rownames("animal_breed_description") |> 
  as.matrix()

# Perform Chi-Squared Test
chi_test <- chisq.test(breed_table, simulate.p.value = TRUE)

# Extract Pearson residuals to examine over/under-representation
residuals <- as.data.frame(chi_test$residuals) |> 
  rownames_to_column("breed") |> 
  pivot_longer(-breed, names_to = "region", values_to = "residual")

# Identify top over-represented breeds per region (Residual > 2)
overrepresented <- residuals |> 
  filter(residual > 2) |> 
  arrange(region, desc(residual))
```

```{r}
# Calculate % above expected

# Extract Observed and Expected values from the test result
observed_counts <- as.data.frame(chi_test$observed)  |> 
  rownames_to_column("breed") |> 
  pivot_longer(-breed, names_to = "region", values_to = "observed")

expected_counts <- as.data.frame(chi_test$expected)  |> 
  rownames_to_column("breed") |> 
  pivot_longer(-breed, names_to = "region", values_to = "expected")

# Join and calculate the percentage difference
comparison_df <- observed_counts |> 
  left_join(expected_counts, by = c("breed", "region")) |> 
  mutate(
    diff_raw = observed - expected,
    percent_diff = (diff_raw / expected) * 100
  )

# View the top over-represented breeds by percentage
breed_over <- comparison_df |> 
  filter(observed > 50)  |>  
  arrange(desc(percent_diff))

write_csv(breed_over, "../data/breed_over.csv")
```

```{r}
# Visualise over-represented breeds in each region

over_breeds <- observed_counts |> 
  left_join(expected_counts, by = c("breed", "region")) |> 
  left_join(residuals, by = c("breed", "region")) |> 
  mutate(
    diff_raw = observed - expected,
    percent_diff = (diff_raw / expected) * 100
  ) |> 
  filter(residual > 2 & observed >= 50) |> 
  group_by(region) |> 
  slice_max(percent_diff, n = 5) |> 
  ungroup()

ggplot(over_breeds, aes(x = reorder_within(breed, percent_diff, region),
                        y = percent_diff, 
                        fill = region)) +
  geom_col() +
  geom_text(aes(label = paste0("+", round(percent_diff), "%")), 
            hjust = -0.1, size = 3.5, fontface = "bold") +
  scale_x_reordered() +
  coord_flip() +
  facet_wrap(~region, scales = "free_y") +
  scale_y_continuous(expand = expansion(mult = c(0, .25))) + 
  labs(
    title = "Regional Dog Breed 'Hotspots'",
    subtitle = "Breeds with the highest % overrepresentation (where Residual > 2)",
    x = NULL,
    y = "% More Common than Expected"
  ) +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold", size = 12),
    panel.grid.minor = element_blank()
  ) +
  guides(fill = "none")
```

Impounds per 1,000 dogs (FY2024 data)

```{r}
regions_impounds <- 
  impounds |>
  filter(impound_date > "2023-07-01") |> 
  left_join(regions_dogs, by = join_by(suburb)) |> 
  group_by(suburb) |> 
  mutate(
    impounds_suburb = n()
  ) |> 
  group_by(local_board) |> 
  mutate(
    impounds_board = n()
  ) |> 
  group_by(region) |> 
  mutate(
    impounds_region = n()
  ) |> 
  ungroup() |> 
  mutate(
    rate_suburb = (impounds_suburb / known_suburb) * 1000,
    rate_board = (impounds_board / known_board) * 1000,
    rate_region = (impounds_region / known_region) * 1000
  ) |> 
  select(suburb, local_board:rate_region) |> 
  unique()

write_csv(regions_impounds, "../data/regions_impounds.csv")
```

Impound rate for attacks/rushing (FY2024 data)

```{r}
regions_impounds_attacks <-
  impounds |>
  filter(impound_date > "2023-07-01") |>
  select(impound_reason, suburb, s_57_type, severity) |> 
  mutate(
    impound_reason = case_match(
      impound_reason,
      c("Relq - Sec 57", "Sec 57") ~ "attack",
      "Sec 57A" ~ "rush", 
      "Sec 15" ~ "neglect",
      "Sec 28" ~ "owner_disqual",
      c("Sec 32", "Sec 33EC") ~ "fail_class",
      "Sec 42" ~ "unreg",
      c("Sec 52 (2a)", "Sec 52 (2b)", "Sec 52A") ~ "roam_control",
      "Sec 56" ~ "barking",
      "Sec 59" ~ "wildlife",
      "Sec 64" ~ "destruct_order",
      "Sec 72A" ~ "seize"
    ) 
  ) |> 
  filter(impound_reason %in% c("attack", "rush")) |> 
  left_join(regions_dogs, by = join_by(suburb)) |> 
  group_by(suburb) |> 
  mutate(
    impounds_attack_suburb = n()
  ) |> 
  group_by(local_board) |> 
  mutate(
    impounds_attack_board = n()
  ) |> 
  group_by(region) |> 
  mutate(
    impounds_attack_region = n()
  ) |> 
  ungroup() |> 
  mutate(
    rate_attack_suburb = (impounds_attack_suburb / known_suburb) * 1000,
    rate_attack_board = (impounds_attack_board / known_board) * 1000,
    rate_attack_region = (impounds_attack_region / known_region) * 1000
  ) |> 
  select(suburb, local_board:rate_attack_region) |> 
  unique()

write_csv(regions_impounds_attacks, "../data/regions_impounds_attacks.csv")
```

Outcome of impound by area

```{r}
regions_impounds_outcome <-
  impounds |> 
  filter(impound_date > "2023-07-01") |>
  select(suburb, status) |> 
  group_by(suburb) |> 
  count(status, name = "count") |> 
  left_join(regions_dogs, by = join_by(suburb)) |>
  mutate(
    status = case_match(
      status, 
      "For Euthanasia" ~ "Euthanised",
      c("Adoption Pending", 
        "Fostered", 
        "Hold - Adoption") ~ "Adopted",
      c("Dead on Arrival",
        "Died in Shelter",
        "Escaped",
        "Hold - General",
        "Hold - Legal Reason",
        "Hold - Temp Test",
        "Processing",
        "Stolen") ~ "Other",
      c("Trfd to Breed Rescue",
        "Trfd to Other TA",
        "Trfd to Rescue Group",
        "Trfd to SPCA") ~ "Rescued",
      .default = status
    ) 
  ) |> 
  group_by(local_board, status) |> 
  summarise(
    count = sum(count)
  ) |> 
  group_by(local_board) |> 
  mutate(
    board_total = sum(count),
    proportion = count / board_total
  )

write_csv(regions_impounds_outcome, "../data/regions_impounds_outcome.csv")

# Visualise euthanasia by local board
# Colour bars by region

regions_impounds_outcome |> 
  filter(status == "Euthanised" & !is.na(local_board)) |> 
  ggplot(aes(x = fct_reorder(local_board, proportion), y = proportion)) +
  geom_col() +
  coord_flip()
```

