---
title: Data Analysis
author: Thomas E. Saunders
---
## Set up

```{r}
#| message: false

library(akldogs)
library(tidyverse)
library(tidytext)
library(fs)
library(knitr)
library(kableExtra)
library(patchwork)
library(scales)
library(Hmisc)
library(corrplot)
library(plotly)

source("../R/functions.R")
source("../R/theme.R")
```

```{r}
#| message: false

report_files <- list.files("../data", pattern = "am-reports", full.names = TRUE)

report_names <- report_files |> 
  path_file() |> 
  path_ext_remove() |> 
  str_extract("[^-]+$")

report_data <- report_files |> 
  set_names(report_names) |> 
  map(read_csv)
```

```{r}
registration <- registration  |> 
  clean_breed_names(starts_with("animal_breed")) |> 
  rename(suburb = suburb_name)

rfs <- rename(rfs, suburb = location_suburb_name)
```

## Auckland Dog Population Overview

```{r}
reg_2025 <-
  registration |> 
  filter(sheet_name == "FY25")
```

As the dog population has increased, rates of registration and desexing have fallen, while rates of menacing dog ownership have increased.

```{r}
#| label: ownership-trends

ownership_trends <-
  report_data$ownership |>
  pivot_wider(
    names_from = category,
    values_from = count
  ) |> 
  arrange(year) |>
  mutate(
    prev_value = lag(dogs),
    change = dogs - prev_value,
    pc_change = change / dogs,
    reg_pc = registered / dogs,
    desex_pc = desexed / dogs,
    men_pc = menacing / dogs
  ) |>
  select(year, dogs, pc_change:men_pc) |> 
  arrange(desc(year)) |> 
  rename(
    "Year" = year,
    "Known Dogs" = dogs,
    "Pop. Change (%)" = pc_change,
    "Registered (%)" = reg_pc,
    "Desexed (%)" = desex_pc,
    "Menacing (%)" = men_pc
  ) |> 
  mutate(
    across(contains("%"),
           ~ scales::percent(., accuracy = 0.1)),
    "Known Dogs" = format(`Known Dogs`, big.mark = ",")
  )

ownership_trends |> 
  kable() |> 
  kable_styling()
```

Top 10 primary breeds as at FY2025:

```{r}
#| label: top-breeds

primary_breed <- reg_2025 |> 
  calc_proportions("animal_breed_description", count_name = "primary_breed")

primary_breed |> 
  rename(
    "Breed" = animal_breed_description,
    "Count" = primary_breed,
    "Proportion" = prop
  ) |> 
  mutate(
    Proportion = scales::percent(Proportion, accuracy = 0.1),
    Count = format(Count, big.mark = ",")
  ) |> 
  head(n = 10) |> 
  kable() |> 
  kable_styling()
```

Dog population secondary breed counts

```{r}
secondary_breed <- reg_2025 |> 
  calc_proportions("animal_breed2_description", count_name = "secondary_breed")

head(secondary_breed, n = 10)
```

Primary breeds most commonly classified as 'menacing' due to behaviour

```{r}
menacing_breed <-
  reg_2025 |>
  filter(classification == "Menacing - observed or reported behaviour of dog") |> 
  count(animal_breed_description, name = "menacing") |>
  left_join(primary_breed |> rename(primary_prop = prop), by = join_by(animal_breed_description)) |>
  filter(primary_breed > 100) |> 
  mutate(menacing_prop = menacing / primary_breed) |> 
  arrange(desc(menacing_prop)) 
  
head(menacing_breed, n = 10)
```

The rate of attacks on people per 1,000 dogs has increased from 6 in FY2016, to 10 in FY2025.

```{r}
attack_rate <-
  report_data$service |> 
  filter(category == "attack-people") |>
  select(-category) |> 
  rename(count_attacks = count) |> 
  left_join(
    report_data$ownership |> 
      filter(category == "dogs") |> 
      select(-category) |> 
      rename(count_dogs = count),
    by = join_by(year)) |> 
  mutate(
    attack_rate = round((count_attacks / count_dogs) * 1000, digits = 1),
  )

head(attack_rate, n = 10)
```

```{r}
#| label: attacks-plot

ggplot(attack_rate |> filter(year > 2015), aes(x = year, y = attack_rate)) +
  geom_line() +
  scale_y_continuous(limits = c(0, 11), breaks = seq(0, 11, by = 1)) +
  scale_x_continuous(breaks = 2016:max(attack_rate$year)) +
  xlab("") +
  ylab("Attacks Per 1,000 Dogs") +
  plot_theme() +
  labs(title = "Auckland Dog Attack Rate Over Time")
```

Impounds

```{r}
#| label: impound-rate

impound_rate <-
  report_data$ownership |> 
  filter(category == "dogs") |> 
  rename("dogs" = "count") |> 
  select(-category) |> 
  left_join(report_data$impounds |> filter(category == "impounds"),
            by = join_by(year)) |> 
  rename("impounds" = "count") |> 
  select(-category) |> 
  mutate(
    impound_rate = round((impounds / dogs) * 1000, digits = 1)
  )

impound_rate |> 
  rename(
    "Year" = year,
    "Known Dogs" = dogs,
    "Impounded Dogs" = impounds,
    "Impound Rate" = impound_rate
  ) |> 
  mutate(
    across(c(`Known Dogs`, `Impounded Dogs`), ~ format(.,big.mark = ","))
  ) |> 
  kable() |> 
  kable_styling()
```


Top 3 primary breeds for each attack severity rating for impounded dogs in FY2024

```{r}
attack_severity <-
  impounds |> 
  filter(impound_date > "2023-07-01" & !is.na(severity)) |> 
  group_by(severity) |> 
  count(primary_breed) |> 
  arrange(desc(severity), desc(n))

attack_severity |> 
  group_by(severity) |>
  slice_max(order_by = n, n = 3) |> 
  arrange(desc(severity))
```

The overall rate of euthanising impounded dogs was dropping between 2014-2019. But after COVID, the rate has increased to be higher than ever before.

```{r}
#| warning: FALSE
#| label: impound-outcome

impound_result <-
  report_data$impounds |> 
  filter(category != "impounds")

impound_result$category <- factor(impound_result$category, 
                                 levels = c("transferred", "returned", "euthanised"))

ggplot(impound_result, aes(x = year, y = count, fill = category)) +
  geom_bar(position = "fill", stat = "identity") +
  scale_fill_manual(values = c("#08519c", "#006d2c", "#a50f15"),
                    labels = tools::toTitleCase) +
  scale_y_continuous(labels = scales::label_percent()) +
  scale_x_continuous(breaks = unique(impound_result$year)) +
  xlab("") +
  ylab("") +
  plot_theme() +
  labs(title = "Outcome For Impounded Dogs Over Time",
       subtitle = "Over half of impounded dogs are now euthanised")
```

Impound outcome by primary breed (population and impound data are both FY2024)

```{r}
#| label: euth-breed

breed_data <- registration |>
  filter(sheet_name == "FY24") |>
  count(primary_breed = animal_breed_description, name = "n_pop") |>
  mutate(prop_pop = n_pop / sum(n_pop)) |>
  inner_join(
    impounds |>
      filter(impound_date > "2023-07-01", status %in% c("Euthanised", "For Euthanasia")) |>
      mutate(primary_breed = str_replace(primary_breed, ".*Pit Bull Terrier.*", "American Pit Bull Terrier")) |>
      count(primary_breed, name = "n_euth") |>
      mutate(prop_euth = n_euth / sum(n_euth)),
    by = "primary_breed"
  ) |>
  mutate(
    diff = prop_euth - prop_pop,
    max_val = pmax(prop_euth, prop_pop) 
  ) |>
  filter(abs(diff) > 0.01) |>
  arrange(abs(diff)) |>
  mutate(primary_breed = fct_inorder(primary_breed))

colours <- c("prop_euth" = "#BF2F24", "prop_pop"  = "#436685")
n_breeds <- nrow(breed_data)

p1 <- breed_data |>
  pivot_longer(cols = starts_with("prop"), names_to = "type", values_to = "val") |>
  ggplot(aes(val, primary_breed)) +
  geom_line(aes(group = primary_breed), color = "#E7E7E7", linewidth = 3.5) +
  geom_point(aes(color = type), size = 3) +
  geom_text(aes(label = percent(val, 0.01), color = type, 
                hjust = if_else(val == max_val, -0.2, 1.2)), size = 3.25) +
  geom_text(data = \(.x) filter(.x, primary_breed == last(primary_breed)),
            aes(label = if_else(type == "prop_pop", "Population", "Euthanised"), color = type),
            nudge_y = 0.8, fontface = "bold", size = 3.25) +
  scale_x_continuous(
  breaks = seq(0, 0.4, by = 0.10),
  labels = \(x) if_else(round(x, 2) %in% round(seq(0, 0.4, 0.1), 2), 
                        percent(x, accuracy = 1), 
                        ""),
  guide = guide_axis(cap = "both")
  ) +
  scale_color_manual(values = colours) +
  coord_cartesian(xlim = c(-0.02, 0.45), ylim = c(1, n_breeds + 1), clip = "off") +
  labs(title = "Breed Profile: Auckland Dog Population vs Euthanised",
       subtitle = "Red 'differences' indicate higher euthanasia than expected", x = NULL, y = NULL) +
  theme_minimal() +
  theme(
    legend.position = "none", 
    panel.grid = element_blank(), 
    plot.title = element_text(face = "bold", hjust = 0.5), 
    plot.subtitle = element_text(hjust = 0.5),
    axis.line.x = element_line(color = "black", linewidth = 0.5),
    axis.ticks.x = element_line(color = "black", linewidth = 0.5),
    axis.ticks.length.x = unit(3, "pt")
    )

p2 <- ggplot(breed_data, aes(x = 0, y = primary_breed)) +
  geom_text(aes(label = percent(diff, 0.1), color = if_else(diff > 0, "prop_euth", "prop_pop")),
            fontface = "bold", size = 3.25) +
  annotate("text", x = 0, y = n_breeds + 0.8, label = "Difference", fontface = "bold", size = 3.25) +
  scale_color_manual(values = colours) +
  coord_cartesian(ylim = c(1, n_breeds + 1), clip = "off") +
  theme_void() +
  theme(legend.position = "none")

p1 + p2 + plot_layout(widths = c(10, 1)) & 
  theme(plot.margin = margin(20, 5, 10, 5))
```

Reasons for euthanasia

```{r}
euthanised_reason <-
  impounds |>
  filter(impound_date > "2023-07-01" & !is.na(euthanasia_reason)) |>
  mutate(
    euthanasia_reason = case_when(
      str_detect(euthanasia_reason, "Health|Parvo") ~ "Health or Parvo Virus",
      str_detect(euthanasia_reason, "Dog vs Dog|Dog vs Domestic|Dog vs Stock|Dog vs Poultry") ~ "S.57 Dog vs Animal",
      .default = euthanasia_reason
    )
  ) |> 
  group_by(euthanasia_reason) |> 
  count(name = "count") |>
  ungroup() |> 
  mutate(
    proportion = count / sum(count)
  ) |> 
  arrange(desc(proportion))

euthanised_reason
```

Total infringements skyrocket in 2025

```{r}
report_data$enforcement |> 
  filter(category == "infringements-total")
```

The bulk of infringements are for failing to register a dog

```{r}
report_data$enforcement |> 
  filter(category == "failure-register")
```

But prosecutions remain at low levels

```{r}
report_data$enforcement |> 
  filter(category == "prosecutions-appeals")
```

## Auckland Dog Population By Region

Known dogs

```{r}
#| label: region-pop

regions_dogs <- 
  registration |>
  filter(sheet_name == "FY25") |> 
  left_join(locality, by = join_by(suburb)) |>
  select(suburb, local_board, region) |> 
  group_by(suburb) |> 
  add_count(suburb, name = "known_suburb") |> 
  group_by(local_board) |> 
  add_count(local_board, name = "known_board") |> 
  group_by(region) |> 
  add_count(region, name = "known_region") |>
  unique() |> 
  arrange(desc(known_suburb)) |> 
  ungroup()

regions_dogs |> 
  select(region, known_region) |> 
  distinct() |> 
  arrange(desc(known_region)) |> 
  rename(
    "Region" = "region",
    "Dogs" = "known_region"
  ) |> 
  mutate(
    Dogs = format(Dogs, big.mark = ",")
  ) |> 
  kable() |> 
  kable_styling()
```

Dog population trends in each Local Board

```{r}
#| label: board-pop

dogs_board <-
  registration |> 
  select(sheet_name, suburb) |> 
  left_join(locality, by = join_by(suburb)) |> 
  group_by(sheet_name, local_board) |> 
  count() |> 
  pivot_wider(names_from = sheet_name, values_from = n) |> 
  mutate(
    change = (FY25 - FY21) / FY21
  ) |> 
  filter(FY25 > 100) |> 
  arrange(desc(change)) |> 
  ungroup()

dogs_board |> 
  mutate(
    across(contains("FY"), comma),
    change = scales::percent(change, accuracy = 0.1)
  ) |> 
  rename(
    "2021" = "FY21",
    "2022" = "FY22",
    "2023" = "FY23",
    "2024" = "FY24",
    "2025" = "FY25",
    "Change" = "change",
    "Local Board" = "local_board"
  ) |> 
  kable() |> 
  kable_styling()
```

```{r}
median(dogs_board$change, na.rm = TRUE)
```

While the median increase across Auckland local boards was 9%, the Interquartile Range (IQR) analysis below identifies Māngere-Ōtāhuhu and Ōtara-Papatoetoe Local Boards as statistical outliers for their significantly higher growth, distinct from the regional trend.

```{r}
# Calculate upper fence
stats <- dogs_board |>
  summarise(
    q1 = quantile(change, 0.25),
    q3 = quantile(change, 0.75),
    iqr = q3 - q1,
    upper_fence = q3 + (1.5 * iqr)
  )

# Flag high outliers
dogs_board_outliers <- dogs_board |>
  mutate(
    is_high_outlier = change > stats$upper_fence,
  ) |>
  filter(is_high_outlier)

dogs_board_outliers
```

Registration data

```{r}
regions_2025 <-
  registration |> 
  filter(sheet_name == "FY25") |> 
  mutate(
    current_reg = case_when(registration_latest_year < 2024 ~ FALSE, .default = TRUE),
    owner_reg = case_when(owner_registration_class == "RDOL" ~ TRUE),
    classification = case_when(classification == "Unknown" ~ FALSE,
                               str_detect(classification, "Menacing|Dangerous") ~ TRUE)
  ) |> 
  left_join(regions_dogs, by = join_by(suburb)) |> 
  select(current_reg, owner_reg, animal_breed_description, animal_desexed, classification, suburb, local_board:known_region) 
  
regions_reg <- 
  regions_2025 |>
  pivot_longer(
    cols = c(current_reg, owner_reg, animal_desexed, classification),
    names_to = "metric",
    values_to = "is_true"
  ) |>
  group_by(suburb, local_board, region, metric) |>
  summarise(
    count = sum(is_true, na.rm = TRUE),
    known_suburb = first(known_suburb),
    known_board = first(known_board),
    known_region = first(known_region),
    .groups = "drop"
  ) |>
  group_by(metric) |>
  mutate(
    p_suburb = count / known_suburb
  ) |>
  group_by(local_board, metric) |>
  mutate(p_board = sum(count) / first(known_board)) |>
  group_by(region, metric) |>
  mutate(p_region = sum(count) / first(known_region)) |>
  ungroup()

write_csv(regions_reg, "../outputs/regions_reg.csv")
```

Compare Local Boards

```{r}
plot_data <- regions_reg |>
  select(local_board, metric, p_board) |>
  distinct() |> 
  mutate(
    metric = metric |> str_replace_all("_", " ") |> str_to_title(),
    local_board = local_board |> str_replace_all("_", " ") |> str_to_title()
  )

plots_list <- plot_data |> group_split(metric)

p <- plots_list |> 
  reduce(\(p_obj, data_slice) {
    p_obj |> 
      add_bars(
        data = data_slice,
        x = ~p_board,
        y = ~fct_reorder(local_board, p_board),
        name = unique(data_slice$metric),
        marker = list(color = region_colours[which(plots_list == list(data_slice))]),
        visible = (identical(data_slice, plots_list[[1]])),
        orientation = 'h',
        showlegend = FALSE,
        hovertemplate = "Board: %{y}<br>Value: %{x:.1%}<extra></extra>"
      )
  }, .init = plot_ly(height = 800))

buttons <- plots_list |> 
  imap(\(data_slice, idx) {
    vis_vector <- rep(FALSE, length(plots_list))
    vis_vector[idx] <- TRUE
    list(
      method = "restyle",
      args = list("visible", as.list(vis_vector)),
      label = unique(data_slice$metric)
    )
  })

interactive_plot <- p |> 
  layout(
    title = list(
      text = "Animal Registration Metrics by Local Board (FY25)",
      x = 0.05,    
      y = 0.98,       
      xanchor = 'left',
      yanchor = 'top',
      font = list(size = 20)
    ),
    margin = list(l = 180, t = 100, r = 30, b = 50),
    xaxis = list(title = "", tickformat = ".0%"),
    yaxis = list(title = ""),
    updatemenus = list(
      list(
        buttons = buttons,
        direction = "down",
        showactive = TRUE,
        x = 0.10, 
        y = 1.05, 
        xanchor = 'left',
        yanchor = 'top',
        bgcolor = "#f8f9fa"
      )
    )
  )

htmlwidgets::saveWidget(interactive_plot, "../outputs/reg_metrics.html", selfcontained = TRUE)
```

The below Spearman rank correlation analysis shows that Local Boards with high current registrations tend to have high rates of desexing and low rates of classified animals, while those with high rates of classifications tend to have lower rates of desexing.

```{r}
#| label: spearman

corr <- plot_data |>
  pivot_wider(names_from = metric, values_from = p_board) |> 
  select(-local_board) |> 
  as.matrix()

cor_results <- rcorr(corr, type = "spearman")

corrplot(cor_results$r, 
         method = "color", 
         type = "upper", 
         addCoef.col = "black", 
         tl.col = "black", 
         diag = FALSE,
         sig.level = 0.05, 
         insig = "blank")
```

Attacks per 1,000 known dogs in each Local Board

```{r}
regions_attacks <- rfs |>
  filter(
    rfs_date >= "2024-07-01",
    str_to_lower(rfs_type) %in% c("dog attack on people", "dog attack on child")
  ) |>
  count(suburb, name = "attacks_suburb") |>
  inner_join(regions_dogs, by = "suburb") |>
  group_by(local_board) |>
  mutate(attacks_board = sum(attacks_suburb)) |>
  group_by(region) |>
  mutate(attacks_region = sum(attacks_suburb)) |>
  ungroup() |>
  mutate(
    rate_suburb = (attacks_suburb / known_suburb) * 1000,
    rate_board  = (attacks_board / known_board) * 1000,
    rate_region = (attacks_region / known_region) * 1000
  ) |>
  select(suburb, local_board, region, starts_with("attacks"), starts_with("known"), starts_with("rate"))

write_csv(regions_attacks, "../outputs/regions_attacks.csv")
```

Primary breed profile for each region

```{r}
regions_breed <-
  registration |> 
  filter(sheet_name == "FY25") |> 
  left_join(locality, by = join_by(suburb)) |> 
  group_by(region) |> 
  count(animal_breed_description, name = "count") |> 
  arrange(desc(count), .by_group = TRUE) |> 
  left_join(regions_dogs |> select(region, known_region) |> unique(), by = join_by(region)) |>
  mutate(
    prop_breed = format(count / known_region, scientific = FALSE)
  )
```

Compare regional breed profiles statistically

```{r}
breeds_region <- regions_breed |>
  perform_chisq(
    group_var = "region", 
    cat_var = "animal_breed_description", 
    count_var = "count"
  )

write_csv(breeds_region, "../outputs/breed_over.csv")
```

```{r}
#| label: region-breeds
#| fig-height: 10

breeds_region |> 
  group_by(group) |> 
  slice_max(percent_diff, n = 5) |> 
  ungroup() |> 
  ggplot(aes(x = reorder_within(category, percent_diff, group), 
             y = percent_diff, 
             fill = group)) +
  geom_col() +
  scale_fill_manual(values = region_colours) +
  geom_text(aes(label = paste0("+", round(percent_diff), "%")), 
            hjust = -0.1, size = 3.5, fontface = "bold") +
  scale_x_reordered() +
  coord_flip() +
  facet_wrap(~group, scales = "free_y", ncol = 1,
             labeller = function(labels) {
      lapply(labels, tools::toTitleCase)
    }) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  plot_theme() +
  theme(legend.position = "none") +
  labs(
    title = "Regional Primary Breed Profiles",
    subtitle = "Top 5 breeds more common than expected",
    x = NULL, y = "% Above Expected"
  )
```

Impounds per 1,000 dogs (FY2024 data)

```{r}
regions_impounds <- impounds |>
  filter(impound_date > "2023-07-01") |>
  count(suburb, name = "impounds_suburb") |>
  inner_join(locality, by = "suburb") |> 
  left_join(regions_dogs |> select(local_board, known_board, known_region), by = "local_board") |>
  unique() |> 
  group_by(local_board) |>
  mutate(impounds_board = sum(impounds_suburb)) |>
  group_by(region) |>
  mutate(impounds_region = sum(impounds_suburb)) |>
  ungroup() |>
  mutate(
    rate_board  = (impounds_board / known_board) * 1000,
    rate_region = (impounds_region / known_region) * 1000
  ) |>
  select(suburb, local_board, region, starts_with("impounds"), starts_with("known"), starts_with("rate"))

write_csv(regions_impounds, "../outputs/regions_impounds.csv")
```

Impound rate for attacks/rushing (FY2024 data)

```{r}
attack_codes <- c("Relq - Sec 57", "Sec 57")
rush_codes   <- c("Sec 57A")

regions_impounds_attacks <- impounds |>
  filter(
    impound_date > "2023-07-01",
    impound_reason %in% c(attack_codes, rush_codes)
  ) |>
  count(suburb, name = "impounds_attack_suburb") |>
  inner_join(locality, by = "suburb") |>
  left_join(regions_dogs |> select(local_board, known_board, known_region), by = "local_board") |> 
  unique() |> 
  group_by(local_board) |>
  mutate(impounds_attack_board = sum(impounds_attack_suburb, na.rm = TRUE)) |>
  group_by(region) |>
  mutate(impounds_attack_region = sum(impounds_attack_suburb, na.rm = TRUE)) |>
  ungroup() |>
  mutate(
    across(starts_with("known"), ~replace_na(.x, 0)),
    rate_attack_board  = (impounds_attack_board / known_board) * 1000,
    rate_attack_region = (impounds_attack_region / known_region) * 1000
  ) |>
  mutate(across(starts_with("rate"), ~if_else(is.infinite(.x), NA_real_, .x))) |>
  select(suburb, local_board, region, starts_with("impounds"), starts_with("known"), starts_with("rate_attack"))

write_csv(regions_impounds_attacks, "../outputs/regions_impounds_attacks.csv")
```

Outcome of impound by area

```{r}
#| label: euth-board

regions_impounds_outcome <- impounds |>
  filter(impound_date > "2023-07-01") |>
  mutate(status = case_match(status, 
      "For Euthanasia" ~ "Euthanised",
      c("Adoption Pending", "Fostered", "Hold - Adoption") ~ "Adopted",
      c("Trfd to Breed Rescue", "Trfd to Other TA", "Trfd to Rescue Group", "Trfd to SPCA") ~ "Rescued",
      c("Dead on Arrival", "Died in Shelter", "Escaped", "Hold - General", 
        "Hold - Legal Reason", "Hold - Temp Test", "Processing", "Stolen") ~ "Other",
      .default = status
    )
  ) |>
  left_join(regions_dogs, by = "suburb") |>
  count(region, local_board, status, name = "count") |>
  group_by(local_board) |>
  mutate(
    board_total = sum(count),
    proportion = count / board_total
  ) |> 
  ungroup()

write_csv(regions_impounds_outcome, "../outputs/regions_impounds_outcome.csv")

regions_impounds_outcome |> 
  filter(status == "Euthanised" & !is.na(local_board)) |> 
  ggplot(aes(x = fct_reorder(local_board, proportion), y = proportion, fill = region)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = region_colours) +
  scale_y_continuous(labels = scales::label_percent()) +
  plot_theme() +
  theme(plot.title = element_text(hjust = 1)) +
  labs(title = "Proportion of Impounded Dogs Euthanised By Local Board",
       subtitle = "") +
  xlab("") +
  ylab("")
```

Impounds by breed

```{r}
#| label: impounds-breed

impounds_breed <- impounds |>
  filter(impound_date > "2023-07-01") |>
  count(primary_breed, name = "impound_count") |>
  inner_join(
    primary_breed |> select(-prop),
    by = join_by(primary_breed == animal_breed_description)
  ) |>
  mutate(impound_rate = round((impound_count / primary_breed.y) * 1000, digits = 1)) |>
  filter(primary_breed.y > 99) |>
  arrange(desc(impound_rate)) |> 
  head(n = 10)

impounds_breed |> 
  rename("Breed" = primary_breed,
         "Impounds" = impound_count,
         "Population" = primary_breed.y,
         "Impound Rate" = impound_rate) |> 
  mutate(
    across(c(`Impounds`, `Population`), ~ format(.,big.mark = ","))
  ) |> 
  kable() |> 
  kable_styling()
```