---
title: Mapping
author: Thomas E. Saunders
---
## Set up

```{r}
#| message: false

library(akldogs)
library(tidyverse)
library(fs)
library(sf)
library(ggnewscale)
library(colorspace)
library(scales)
library(patchwork)

source("../R/functions.R")
source("../R/theme.R")
```

Read in regional data created in Data Analysis notebook

```{r}
#| message: false

regional_files <- list.files(path = "../outputs", 
                           pattern = "regions|breed", 
                           full.names = TRUE)

regional_data <- 
  regional_files |> 
  map(read_csv) |> 
  set_names(path_file(str_extract(regional_files, "[^-]+(?=\\.[^.]+$)")) |>
              path_ext_remove())
```

Get Auckland human population by Local Board

```{r}
#| message: false

# Source: https://figure.nz/chart/bSr6yQmn9V9BFrXK

auck_pop <- 
  read_csv("../data/2023-auckland-population.csv") |> 
  filter(`Census year` == "2023" & `Local board area` != "Auckland") |> 
  mutate(
    local_board = paste0(`Local board area`, " Local Board Area"),
    local_board = case_when(
      local_board == "Auckland Local Board Area" ~ "Total Auckland",
      .default = local_board)
  ) |> 
  rename(census_year = `Census year`,
         population = Value) |> 
  select(census_year, local_board, population)
```

Import NZ map shapefile with local board divisions

```{r}
#| message: false

# Source: https://datafinder.stats.govt.nz/layer/120947-territorial-authority-local-board-2025-clipped/

shape_nz <- st_read("../data/shapefiles/nz/territorial-authority-local-board-2025-clipped.shp") |> 
  rename(local_board = TALB2025_1) |> 
  select(local_board, geometry) |> 
  filter(local_board != c("Area Outside Territorial Authority", "Chatham Islands Territory"))
```

Join locality data to shapefile data

```{r}
auckland <-
  shape_nz |>
  st_simplify(dTolerance = 100) |> 
  mutate(
    include = case_when(local_board %in% auck_pop$local_board ~ TRUE,
                        local_board %in% c("Kaipara District", 
                                           "Thames-Coromandel District",
                                           "Waikato District",
                                           "Hauraki District",
                                           "Whangarei District") ~ TRUE)
  ) |>
  left_join(akldogs::locality |> 
              select(local_board, region) |>
              unique(),
            by = join_by(local_board)) |>
  mutate(
    region = case_when(local_board == "Aotea/Great Barrier Local Board Area" ~ "north",
                       .default = region)
  ) |> 
  filter(include == TRUE)

auckland$region <- as.factor(auckland$region)
```

## Maps

Map of NZ with border around Auckland

```{r}
map_nz <- shape_nz |> 
  st_simplify(dTolerance = 100) |>
  ggplot() +
  geom_sf(fill = "#b0bc6d", colour = "black") +
  geom_rect(
    xmin = 1700000,
    ymin = 5869755,
    xmax = 1830000,
    ymax = 6010000,
    fill = NA, 
    colour = "darkred",
    linewidth = 1
  ) +
  theme_void() +
  theme(legend.position = "none",
        panel.background = element_rect(fill = "lightblue"))

map_nz
```

Auckland regions

```{r}
map_regions <- ggplot() +
  geom_sf(data = auckland, colour = NA) +
  geom_sf(data = auckland |> 
            filter(!is.na(region)), aes(fill = region), colour = NA) + 
  scale_fill_manual(values = region_colours) +
  coord_sf(
    xlim = c(1700000, 1830000),
    ylim = c(5869755, 6010000),
    expand = FALSE
  ) +
  theme_void() +
  theme(legend.position = "none")

map_regions
```

Auckland Local Boards

```{r}
map_boards <- ggplot() +
  geom_sf(data = auckland, aes(fill = local_board), colour = NA) + 
  scale_fill_manual(values = board_colours) +
  coord_sf(
    xlim = c(1700000, 1830000),
    ylim = c(5869755, 6010000),
    expand = FALSE
  ) +
  theme_void() +
  theme(legend.position = "none")

map_boards
```

Join maps together

```{r}
#| label: map-regions-boards
#| fig-cap: "Left: Auckland regions (North = blue, West = red, Central = green, East = orange, South = purple). Right: Auckland's 21 Local Boards, coloured based on their parent region."

map_regions + map_boards
```

Proportion of desexed dogs by Local Board

```{r}
#| label: map-desex

map_desex <-
  auckland |> 
  left_join(regional_data$regions_reg |> filter(metric == "animal_desexed"), 
            by = join_by(local_board)) |> 
  select(local_board, p_board, geometry) |> 
  unique() |> 
  filter(!is.na(p_board))

plot_local_board(auckland, 
                 map_desex,
                 p_board,
                 "Desexed Dogs",
                 "Proportion of dogs in each Local Board Area") +
  scale_fill_gradient(low = "red",
                      high = "green",
                      name = "Desexed",
                      labels = label_percent())
```

Proportion of dogs with a classification under the Dog Control Act by Local Board

```{r}
#| label: map-class

map_class <-
  auckland |> 
  left_join(regional_data$regions_reg |> filter(metric == "classification"), 
            by = join_by(local_board)) |> 
  select(local_board, p_board, geometry) |> 
  unique() |> 
  filter(!is.na(p_board))

plot_local_board(auckland, 
                 map_class,
                 p_board,
                 "Menacing & Dangerous Dogs",
                 "Proportion of dogs in each Local Board Area",
                 "Classifications",
                 label_percent())
```

Rate of dog attacks by Local Board

```{r}
#| label: map-attacks

map_attacks <-
  auckland |> 
  left_join(regional_data$regions_attacks |> 
              select(local_board, rate_board) |> 
              unique(),
            by = join_by(local_board)) |> 
  select(local_board, rate_board, geometry) |> 
  filter(!is.na(rate_board))

plot_local_board(auckland, 
                 map_attacks,
                 rate_board,
                 "Dog Attacks",
                 "Per 1,000 dogs in each Local Board Area",
                 "Attack Rate")
```

Rate of dog impounds by Local Board

```{r}
#| label: map-impounds

map_impounds <-
  auckland |> 
  left_join(regional_data$regions_impounds, by = join_by(local_board)) |> 
  select(local_board, rate_board, geometry) |> 
  unique() |> 
  filter(!is.na(rate_board))

plot_local_board(auckland, 
                 map_impounds,
                 rate_board,
                 "Dog Impounds",
                 "Per 1,000 dogs in each Local Board Area",
                 "Impound Rate")
```

Rate of dog impounds due to attacks on people by Local Board

```{r}
map_impounds_attacks <-
  auckland |> 
  left_join(regional_data$regions_impounds_attacks, by = join_by(local_board)) |> 
  select(local_board, rate_attack_board, geometry) |> 
  unique() |> 
  filter(!is.na(rate_attack_board))

plot_local_board(auckland, 
                 map_impounds_attacks,
                 rate_attack_board,
                 "Dogs Impounded For Attacks",
                 "Per 1,000 dogs in each Local Board Area",
                 "Impound Rate")
```

Proportion of impounded dogs euthanised by Local Board

```{r}
#| label: map-euth

map_impounds_euthanised <-
  auckland |> 
  left_join(regional_data$regions_impounds_outcome, by = join_by(local_board)) |> 
  select(local_board, status, proportion, geometry) |> 
  unique() |> 
  filter(status == "Euthanised" & local_board != "Aotea/Great Barrier Local Board Area")

plot_local_board(auckland, 
                 map_impounds_euthanised,
                 proportion,
                 "Euthanasia of Impounded Dogs",
                 "Over half put down in some southern areas",
                 "Euthanised",
                 label_percent())
```

